Effectful Semantics for Substructural SSA

### 2024-03-11 syco12/paper.tex
Tap

\documentclass[acmsmall,screen,review]{acmart}

@@

\title{Effectful Semantics for Substructural SSA}
% Or: SSA is Freyd Categories

@@

Unfortunately, substitution for blocks is a little more complicated: the naive
choice of attempting to show \(\haslb{\Gamma}{t}{\ms{L}} \implies
\haslb{\Theta}{[\gamma]t}{\ms{L}}\) does not work, since \(\ms{L}\) may require
variables from \(\Gamma\) to be live which simply do not exist in \(\Theta\). To
solve this problem, we need to bind the terms in \(\gamma\) to variables with
the correct names for the live variable sets of desired targets. To do this, we
introduce the concept of a \textit{label-substitution} \(\mc{L}\), which we
define as a function that takes labels \(\lbl{\ell}\) and variables \(x\) to
blocks \(t\) such that, for any fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell},
x) = \mc{L}(\lbl{\ell}, a)\). We introduce the judgement
\(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to mean that for each
\(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have that \(\mc{L}(\lbl{\ell}, x)\)
targets \(\ms{K}\). Formally, \rle{lb-subst-nil} states that this is vacuously
true for \(\ms{L} = \cdot\), while \rle{lb-subst-cons} says that if
\(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
\thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{(\ms{L},
\lhyp{\lbl{\ell}}{\Gamma}{A})}{\ms{K}}\). Given a context \(\gamma\) and a
label-set \(\ms{L}\), we can define an associated label-substitution
\(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
\ref{fig:ssa-subst-ops}: using this, we can state the full substitution lemma as
follows:
\begin{lemma}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then
    \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) and
    \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}\)
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such
    that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\) and
    \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
    \ms{W}'\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
    \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\), and
    \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
  \end{itemize}
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), we have that for all
  \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
\end{lemma}

@@

We can now state the theorem of semantic substitution, which essentially states
that the semantics of a substituted term or block is the same as the semantics
of that term or block preceded by let-bindings corresponding to the variables in
the substitution, which is the same as the semantics of the original term or
block precomposed with the semantics of the substitution. Similarly, the
semantics of a \textit{label-substituted} block is the same as the semantics of
the original block \textit{postcomposed} with the semantics of that
label-substitution. Stated formally, we obtain:
\begin{theorem}[Semantic Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
    \(
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
      = \dnt{\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}}
      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
    \)
    \item For all 
      \(\haslb{\Gamma}{t}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      = \dnt{\haslb{\Theta}{\exprletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
    \]
    \item For all 
      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
      \[
        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
      \]
  \end{itemize}  
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
  \begin{itemize} 
    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
    \(
      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
    \)
    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
    \[
      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
    \]
    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
  \end{itemize}
\end{theorem}

@@

\section{Proofs}

\begin{lemma} We have that
  \begin{enumerate}
    \item Splitting commutes: if \(\csplits{\Gamma}{\Delta}{\Xi}\) then \(\csplits{\Gamma}{\Xi}{\Delta}\), and \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}} = \dnt{\csplits{\Gamma}{\Xi}{\Delta}};\sigma\)
    \item Splitting associates: if \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\csplits{\Xi}{\Theta}{\Phi}\) then there exists \(\Gamma_{\Delta, \Theta}\) such that \(\csplits{\Gamma}{\Gamma_{\Delta, \Theta}}{\Phi}\) and \(\csplits{\Gamma_{\Delta, \Theta}}{\Delta}{\Theta}\), and, for \emph{all} such \(\Gamma_{\Delta, \Theta}\), we have \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}};\dnt{\Delta} \otimes \csplits{\Xi}{\Theta}{\Phi} = \dnt{\csplits{\Gamma}{\Gamma_{\Delta, \Theta}}{\Phi}};\dnt{\csplits{\Gamma_{\Delta, \Theta}}{\Delta}{\Theta}} \otimes \dnt{\Phi};\alpha\)
    \item Weakening composes: if \(\cwk{\Gamma}{\Delta}\) and \(\cwk{\Delta}{\Xi}\), then \(\cwk{\Gamma}{\Xi}\)
    \item Splitting and weakening compose on the right: if \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\cwk{\Delta}{\Delta'}\) then \(\csplits{\Gamma}{\Delta'}{\Xi}\), furthermore,
    \(\dnt{\csplits{\Gamma}{\Delta'}{\Xi}} = \dnt{\csplits{\Gamma}{\Delta}{\Xi}};\dnt{\cwk{\Delta}{\Delta'}} \otimes \dnt{\Xi}\)
    \item Splitting and weakening compose on the left: if \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\cwk{\Gamma'}{\Gamma}\) then \(\csplits{\Gamma'}{\Delta'}{\Xi}\), furthermore,
    \(\dnt{\csplits{\Gamma'}{\Delta}{\Xi}} = \dnt{\cwk{\Gamma'}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\)
  \end{enumerate}
\end{lemma}

\begin{proof}
  By induction.
\end{proof}

\begin{lemma}[Weakening] \
  \begin{enumerate}
    \item If \(\cwk{\Gamma}{\Delta}\) and \(q \subseteq r\),
    \(\hasty{\Delta}{p}{a}{A}{r}\), then \(\hasty{\Gamma}{p}{a}{A}{q}\)
    \item If \(\cwk{\Gamma}{\Delta}\), \(\haslb{\Delta}{t}{\ms{L}}\), then
    \(\haslb{\Gamma}{t}{\ms{L}}\)
    \item If \(\lwk{\ms{L}}{\ms{K}}\), \(\haslb{\Gamma}{t}{\ms{L}}\) then
    \(\haslb{\Gamma}{t}{\ms{K}}\)
  \end{enumerate}
\end{lemma}

\begin{proof}
  By induction.
\end{proof}

\begin{lemma}[Substitution Splitting] 
  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\),
  then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that
  \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\),
  \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), and
  \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). In particular, if
  \(\cwk{\Gamma}{\Delta}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then
  \(\issubst{\gamma}{\Theta}{\Delta}\).
\end{lemma}

\begin{proof} 
  We proceed by induction on derivations \(\csplits{\Gamma}{\Delta}{\Xi}\). In
  the \rle{split-nil} case, we have that \(\issubst{\gamma}{\cdot}{\cdot}\) by
  \rle{subst-nil}. For all other cases, since \(\issubst{\gamma}{\Theta}{\Gamma,
  \thyp{x}{A}{q}}\), we have by \rle{subst-cons} that
  \(\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}\),
  \(\issubst{\gamma}{\Theta_\Gamma}{\Gamma}\),
  \(\hasty{\Theta_\Gamma}{1}{\gamma(x)}{A}{q}\). Since
  \(\csplits{\Gamma}{\Delta}{\Xi}\), by induction, we have that
  \(\csplits{\Theta_\Gamma}{\Theta_\Delta}{\Theta_\Xi}\) where \(\Theta_\Delta\)
  and \(\Theta_\Xi\) have the desired properties above.
  \begin{itemize}
    \item \rle{split-left}: by left split association, we have that there exists
    \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\csplits{\Theta}{\Theta_{\Delta,
    \thyp{x}{A}{q}}}{\Theta_\Xi}\), \(\csplits{\Theta_{\Delta,
    \thyp{x}{A}{q}}}{\Theta_\Delta}{\Theta_x}\). Using the fact that
    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), we can conclude by
    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Delta,
    \thyp{x}{A}{q}}}{\Delta, \thyp{x}{A}{q}}\); the desired result hence
    follows.
    \item \rle{split-right}: by right split association, we have that there
    exists \(\Theta_{\Xi, \thyp{x}{A}{q}}\),
    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_{\Xi, \thyp{x}{A}{q}}}\),
    \(\csplits{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Theta_\Xi}{\Theta_x}\). Using the
    fact that \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by
    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
    \thyp{x}{A}{q}}\); the desired result hence follows.
    \item \rle{split-dup}: since \(\tlin{A}{\trel}\), it follows by
    \rle{subst-cons} that \(\tlin{\Theta_x}{\trel}\), and hence that
    \(\csplits{\Theta_x}{\Theta_x}{\Theta_x}\). Therefore, by split association,
    we have that there exist \(\Theta_{\Delta, \thyp{x}{A}{q}}\), \(\Theta_{\Xi,
    \thyp{x}{A}{q}}\) such that \(\csplits{\Theta}{\Theta_{\Delta,
    \thyp{x}{A}{q}}}{\Theta_{\Xi, \thyp{x}{A}{q}}}\). Using the facts that
    \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\),
    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), we can conclude by \rle{subst-cons}
    that \(\issubst{\gamma}{\Theta_{\Delta, \thyp{x}{A}{q}}}{\Delta,
    \thyp{x}{A}{q}}\), \(\issubst{\gamma}{\Theta_{\Xi, \thyp{x}{A}{q}}}{\Xi,
    \thyp{x}{A}{q}}\); the desired result hence follows.
    \item \rle{split-drop}: since \(\tlin{A}{\taff}\), it follows by
    \rle{subst-cons} that \(\tlin{\Theta_x}{\taff}\), and hence that
    \(\cwk{\Theta_x}{\cdot}\). Therefore, by split association,
    \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\), giving the desired result.
  \end{itemize}
\end{proof}

\begin{lemma}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{enumerate}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) \label{lemma:substitution:term}
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
    \label{lemma:substitution:block}
    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
    \ms{W}'\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
    \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\), and
    \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
    \label{lemma:substitution:cfg}
  \end{enumerate}
\end{lemma}

\begin{proof} 
  We begin by proving \ref{lemma:substitution:term} by induction on derivations \(\hasty{\Gamma}{p}{a}{A}{q}\):
  \begin{itemize}
    \item \rle{var}: we have that \(\cwk{\Gamma}{\thyp{x}{A}{q}}\), and
    therefore that \(\issubst{\gamma}{\Theta}{\thyp{x}{A}{q}}\); by
    \rle{subst-cons}, we obtain the desired result that
    \(\hasty{\Theta}{p}{\gamma(x)}{A}{q}\).
    \item \rle{app}: by induction, we have that
    \(\hasty{\Theta}{1}{[\gamma]a}{A}{q}\), and therefore we can apply \rle{app}
    to obtain \(\hasty{\Theta}{p}{[\gamma](f\;a)}{B}{q}\) as desired.
    \item \rle{pair}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\) and
    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Therefore, by induction, we have that
    \(\hasty{\Theta_\Delta}{1}{[\gamma]a}{A}{q}\) and
    \(\hasty{\Theta_\Xi}{1}{[\gamma]b}{B}{q}\); we can simply apply \rle{pair}
    to obtain our desired result that \(\hasty{\Theta}{p}{[\gamma](a, b)}{A
    \otimes B}{q}\)
    \item \rle{true}, \rle{unit}, \rle{false}: these cases hold trivially since
    \(\cwk{\Theta}{\cdot}\)
    \item \rle{let}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
    and \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Taking \(x\) to be a fresh
    variable (with \(\gamma(x) = x\)), we can conclude by \rle{subst-cons} that
    \(\issubst{\gamma}{\Theta_\Delta, \thyp{x}{A}{}}{\Delta, \thyp{x}{A}{}}\).
    Therefore, by induction, we have that \(\hasty{\Theta_\Delta, \thyp{x}{A}{}}{p}{[\gamma]e}{B}{q}\) and \(\hasty{\Theta_\Xi}{1}{[\gamma]a}{A}{q}\); we
    can hence simply apply \rle{let} to obtain
    \(\hasty{\Theta}{p}{[\gamma]\letexpr{x}{a}{e}}{B}{q}\), as desired.
    \item \rle{let2}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
    and \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Taking \(x, y\) to be fresh
    variables (with \(\gamma(x) = x\), \(\gamma(y) = y\)), we can conclude by
    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_\Delta, \thyp{x}{A}{},
    \thyp{y}{B}{}}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}\). Therefore, by
    induction, we have that \(\hasty{\Theta_\Delta, \thyp{x}{A}{},
    \thyp{y}{B}{}}{p}{[\gamma]e}{C}{q}\) and
    \(\hasty{\Theta_\Xi}{1}{[\gamma]a}{A \otimes B}{q}\); we can hence simply
    apply \rle{let2} to obtain
    \(\hasty{\Theta}{p}{[\gamma]\letexpr{(x, y)}{a}{e}}{C}{q}\), as desired.
  \end{itemize}
  Using \ref{lemma:substitution:term}; we can now prove
  \ref{lemma:substitution:block} and \ref{lemma:substitution:cfg} by mutual
  induction on derivations as follows:
  \begin{itemize}
    \item \rle{br}: \todo{this}
    \item \rle{let-blk}: \todo{this}
    \item \rle{ite}: \todo{this}
    \item \rle{let2-blk}: \todo{this}
    \item \rle{where}: \todo{this}
    \item \rle{nil-br}: \todo{this}
    \item \rle{cons-br}: \todo{this}
  \end{itemize}
\end{proof}

\begin{lemma}[Label Substitution] 
  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
  \begin{itemize}
    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\), \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
  \end{itemize}
\end{lemma}

\begin{proof} 
  We proceed by mutual induction on derivations as follows:
  \begin{itemize}
    \item \rle{br}: \todo{this}
    \item \rle{let-blk}: \todo{this}
    \item \rle{ite}: \todo{this}
    \item \rle{let2-blk}: \todo{this}
    \item \rle{where}: \todo{this}
    \item \rle{nil-br}: \todo{this}
    \item \rle{cons-br}: \todo{this}
  \end{itemize}
\end{proof}

\begin{lemma}[Semantic Upgrade]
  If \(\hasty{\Gamma}{1}{a}{A}{q}\), we have that
  \(\dnt{\hasty{\Gamma}{0}{a}{A}{q}} =
  \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\)
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

\begin{lemma}[Semantic Weakening] 
  If \(\cwk{\Gamma}{\Delta}\), then 
  \begin{itemize}
    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
    \(
      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
    \)
    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
    \(
      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
    \)
  \end{itemize}

  Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
  \(
    \dnt{\haslb{\Gamma}{t}{\ms{L}}}
    ; \dnt{\lwk{\ms{L}}{\ms{K}}}
    = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
  \)
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

\begin{lemma}[Substitution Coherence] \
  \begin{itemize}
    \item Given any two derivations \(D_1: \issubst{\gamma}{\Theta}{\Gamma}\), \(D_2: \issubst{\gamma'}{\Theta}{\Gamma}\) such that \(\forall x \in \Theta, \gamma(x) = \gamma'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1: \lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), \(D_2: \lbsubst{\mc{L}'}{\ms{L}}{\ms{K}}\) such that \(\forall \lbl{\ell} \in \ms{L}, \mc{L}(x) = \mc{L}'(x)\), we have \(\dnt{D_1} = \dnt{D_2}\)
  \end{itemize}
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

\begin{lemma}[Semantic Substitution Splitting] 
  Given \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Delta}{\Xi}\),
  if \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\) and \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\), \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\), then
  \[
    \dnt{\issubst{\gamma}{\Theta}{\Gamma}};\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
    = \dnt{\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}};
      \dnt{\issubst{\gamma}{\Theta_\Delta}{\Delta}}
      \otimes \dnt{\issubst{\gamma}{\Theta_\Xi}{\Xi}}
  \]
\end{lemma}

\begin{proof} \
  \TODO{this}
\end{proof}

>>>>

\documentclass[acmsmall,screen]{acmart}

@@

\title{Effectful Semantics for Substructural SSA}

@@

Unfortunately, substitution for blocks is a little more complicated: the naive
choice of attempting to show \(\haslb{\Gamma}{t}{\ms{L}} \implies
\haslb{\Theta}{[\gamma]t}{\ms{L}}\) does not work, since \(\ms{L}\) may require
variables from \(\Gamma\) to be live which simply do not exist in \(\Theta\). To
solve this problem, we need to bind the terms in \(\gamma\) to variables with
the correct names for the live variable sets of desired targets. To do this, we
introduce the concept of a \textit{label-substitution} \(\mc{L}\), which we
define as a function that takes labels \(\lbl{\ell}\) and variables \(x\) to
blocks \(t\) such that, for any fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell},
x) = \mc{L}(\lbl{\ell}, a)\). We introduce the judgement
\(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to mean that for each
\(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have that \(\mc{L}(\lbl{\ell}, x)\)
targets \(\ms{K}\). Formally, \rle{lb-subst-nil} states that this is vacuously
true for \(\ms{L} = \cdot\), while \rle{lb-subst-cons} says that if
\(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
\thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{(\ms{L},
\lhyp{\lbl{\ell}}{\Gamma}{A})}{\ms{K}}\). Given a context \(\gamma\) and a
label-set \(\ms{L}\), we can define an associated label-substitution
\(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
\ref{fig:ssa-subst-ops}: using this, we can state the full substitution lemma as
follows:
\begin{lemma}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then
    \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) and
    \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}\)
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such
    that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\) and
    \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
    % \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
    % \ms{W}'\) such that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
    % \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\), and
    % \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
  \end{itemize}
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), we have that for all
  \(\haslb{\Gamma}{t}{\ms{L}}\), \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
\end{lemma}

@@

We can now state the theorem of semantic substitution, which essentially states
that the semantics of a substituted term or block is the same as the semantics
of that term or block preceded by let-bindings corresponding to the variables in
the substitution, which is the same as the semantics of the original term or
block precomposed with the semantics of the substitution. Similarly, the
semantics of a \textit{label-substituted} block is the same as the semantics of
the original block \textit{postcomposed} with the semantics of that
label-substitution. Stated formally, we obtain:
\begin{theorem}[Semantic Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
    \(
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
      = \dnt{\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}}
      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
    \)
    \item For all 
      \(\haslb{\Gamma}{t}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      = \dnt{\haslb{\Theta}{\exprletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
    \]
    % \item For all 
    %   \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
    %   \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
    %   \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
    %   \[
    %     \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
    %     ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
    %     = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
    %     ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
    %     + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
    %   \]
  \end{itemize}  
  % Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
  % \begin{itemize} 
  %   \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
  %   \(
  %     \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
  %     = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
  %   \)
  %   \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
  %   \[
  %     \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
  %     + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
  %     = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
  %   \]
  %   where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
  % \end{itemize}
\end{theorem}

@@
